
# Module 1 Exercise Results

Load packages and data, subset needed columns:
```{r, class.source="codeblock",eval=TRUE}
library(MASS)
data(crabs)
```

Learn more about the data:
```{r, class.source="codeblock",eval=TRUE}
?crabs
head(crabs)
```

Subset needed columns:
```{r, class.source="codeblock",eval=TRUE}
crabs_meas <-  crabs[,c("FL","RW","CL","CW","BD")]
```

Perform hierarchical clustering:

```{r, class.source="codeblock",eval=TRUE}
c_dist <- dist(crabs_meas)
c_hclust <- hclust(c_dist)
plot(c_hclust)
```

Colour-code samples based on cluster assignment. Assume there are two clusters.

```{r, class.source="codeblock",eval=TRUE}
c_clusters = cutree(c_hclust,k = 2)
```

Now create a pairs plot, but colour-code by: 
1. by gene-expression based clusters
2. by species
3. by sex

```{r, class.source="codeblock",eval=TRUE}
pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[c_clusters]
)

pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[factor(crabs$sp)]
)

pairs(
    crabs_meas, 
    col = c("orchid","forestgreen")[factor(crabs$sex)]
)
```

**ADD PIECE ABOUT MISSINGNESS**

### Module 2 exercise

PCA
```{r, class.source="codeblock",eval=TRUE}
c_pcs = prcomp(crabs_meas)
```

Plot PC projections (embeddings). 
```{r, class.source="codeblock",eval=TRUE}
pairs(c_pcs$x, col = c("orchid","forestgreen")[factor(crabs$sp)])
pairs(c_pcs$x, col = c("orchid","forestgreen")[factor(crabs$sex)])
```

tSNE:

```{r, class.source="codeblock",eval=TRUE}
library(tsne)
c_tsne10 = tsne(crabs_meas,perplexity = 10)
c_tsne20 = tsne(crabs_meas,perplexity = 20)
c_tsne50 = tsne(crabs_meas,perplexity = 50)
c_tsne100 = tsne(crabs_meas,perplexity = 100)
```


sex_cols = c("orchid","forestgreen")[factor(crabs$sex)]

Color-code tSNE plot by species, try various perplexity levels:
```{r, class.source="codeblock",eval=TRUE}
species_cols = c("orchid","forestgreen")[factor(crabs$sp)]
par(mfrow=c(2,2))
plot(c_tsne10[,1],
     c_tsne10[,2],
     main = "Perplexity = 10",
     col = species_cols)

plot(c_tsne20[,1],
     c_tsne20[,2],
     main = "Perplexity = 20",
     col = species_cols)
plot(c_tsne50[,1],
     c_tsne50[,2],
     main = "Perplexity = 50",
     col = species_cols)
plot(c_tsne100[,1],
     c_tsne100[,2],
     main = "Perplexity = 100",
     col = species_cols)
```

Now do the same, but colour-code for sex:

```{r, class.source="codeblock",eval=TRUE}
sex_cols = c("orchid","forestgreen")[factor(crabs$sex)]
par(mfrow=c(2,2))
plot(c_tsne10[,1],
     c_tsne10[,2],
     main = "Perplexity = 10",
     col = sex_cols)
plot(c_tsne20[,1],
     c_tsne20[,2],
     main = "Perplexity = 20",
     col = sex_cols)
plot(c_tsne50[,1],
     c_tsne50[,2],
     main = "Perplexity = 50",
     col = sex_cols)
plot(c_tsne100[,1],
     c_tsne100[,2],
     main = "Perplexity = 100",
     col = sex_cols)
```


Run UMAP
```{r, class.source="codeblock",eval=TRUE}
library(umap)
c_umap <- umap(crabs_meas)
str(c_umap)

par(mfrow=c(1,2))
plot(c_umap$layout[,1],
     c_umap$layout[,2],
     col = species_cols, pch = 19, 
     main = "Colored by species")

plot(c_umap$layout[,1],
     c_umap$layout[,2],
     col = sex_cols, pch = 19, 
     main = "Colored by sex")
```
